\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-12 paper.qmd
Updated ordering of points symbols

DHC: Will need to have more bins in each of the methods to illustrate this properly.

To illustrate a difference in how the methods organise points in the NLDR layout, simulated 4D data having five Gaussian clusters is used. @fig-five-gau-projs a1, b1, c1 show the 2D layouts for (a) t-SNE, (b) UMAP, and (c) PaCMAP, respectively. The default hyper-parameters are used. All three methods show the five clusters, with varying degrees of separation.

The models are fitted to these layouts, but we focus on a single cluster to illustrate the curious detail. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the 4D space. In this projection the difference between methods can be seen. These clusters are fully 4D in nature, so we would expect the model to be a *crumpled 2D sheet* that stretches in all four dimensions. This is what is observed for t-SNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape! What this means is that there has to be some ordering of points in the 2D PaCMAP layout that induces the flat model, likely that the points are organised by the global principal components. So the layout of the model in 4D doesn't reflect the dimension of each cluster. DHC: is there something in the documentation that might help to explain what has happened?

This pattern can also be observed by examining the error from the model fit (@fig-five-gau-projs a3, b3, c3). With t-SNE and UMAP the error is relatively uniformly distributed, but with PaCMAP there is more error in the center of the 2D layout, reflecting that there are more points in the middle that are further from the fitted model.

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-height: 10
#| fig-width: 10
#| fig-pos: H
#| fig-cap: "The sparse end of the structure exhibits a high model error, while the dense end shows a low model error. The tSNE layouts of the C-shaped cluster, with uniform points (a1) and dense points (b1), along with their selected projections (a2 and b2), are displayed. Darker colors represent a high error, while lighter colors indicate a low error. The uniformly distributed points demonstrate random model errors throughout the structure, whereas the dense structure shows large model error at the sparse end."

>>>>

<!-- DHC: Will need to have more bins in each of the methods to illustrate this properly. -->

To illustrate a difference in how the methods organise points in the NLDR layout, simulated $4\text{-}D$ data having five Gaussian clusters is used. @fig-five-gau-projs a1, b1, c1 show the \gD{} layouts for (a) tSNE, (b) UMAP, and (c) PaCMAP, respectively. The default hyper-parameters are used. All three methods show the five clusters, with varying degrees of separation.

The models are fitted to these layouts, but we focus on a single cluster to illustrate the curious detail. @fig-five-gau-projs a2, b2, c2 show the fitted models in a projection of the $4\text{-}D$ space. In this projection the difference between methods can be seen. These clusters are fully $4\text{-}D$ in nature, so we would expect the model to be a *crumpled \gD{} sheet* that stretches in all four dimensions. This is what is observed for tSNE and UMAP. The curious detail is that the model for PaCMAP is closer to a *pancake* in shape! What this means is that there has to be some ordering of points in the \gD{} PaCMAP layout that induces the flat model, likely that the points are organised by the global principal components. So the layout of the model in $4\text{-}D$ doesn't reflect the dimension of each cluster. DHC: is there something in the documentation that might help to explain what has happened?

This pattern can also be observed by examining the error from the model fit (@fig-five-gau-projs a3, b3, c3). With tSNE and UMAP the error is relatively uniformly distributed, but with PaCMAP there is more error in the center of the \gD{} layout, reflecting that there are more points in the middle that are further from the fitted model.

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-height: 10
#| fig-width: 10
#| fig-pos: H
#| fig-cap: "The sparse end of the structure exhibits a high model error, while the dense end shows a low model error. The tSNE layouts of the C-shaped cluster, with uniform points (a1) and dense points (b1) are colored according to their high-dimensional error, along with their selected projections (a2 and b2), are displayed. Darker colors represent a high error, while lighter colors indicate a low error. The uniformly distributed points demonstrate random model errors throughout the structure, whereas the dense structure shows large model error at the sparse end."

### 2025-01-12 paper.tex
Updated ordering of points symbols

DHC: Will need to have more bins in each of the methods to illustrate
this properly.

To illustrate a difference in how the methods organise points in the
NLDR layout, simulated 4D data having five Gaussian clusters is used.
Figure~\ref{fig-five-gau-projs} a1, b1, c1 show the 2D layouts for (a)
t-SNE, (b) UMAP, and (c) PaCMAP, respectively. The default
hyper-parameters are used. All three methods show the five clusters,
with varying degrees of separation.

The models are fitted to these layouts, but we focus on a single cluster
to illustrate the curious detail. Figure~\ref{fig-five-gau-projs} a2,
b2, c2 show the fitted models in a projection of the 4D space. In this
projection the difference between methods can be seen. These clusters
are fully 4D in nature, so we would expect the model to be a
\emph{crumpled 2D sheet} that stretches in all four dimensions. This is
what is observed for t-SNE and UMAP. The curious detail is that the
model for PaCMAP is closer to a \emph{pancake} in shape! What this means
is that there has to be some ordering of points in the 2D PaCMAP layout
that induces the flat model, likely that the points are organised by the
global principal components. So the layout of the model in 4D doesn't
reflect the dimension of each cluster. DHC: is there something in the
documentation that might help to explain what has happened?

This pattern can also be observed by examining the error from the model
fit (Figure~\ref{fig-five-gau-projs} a3, b3, c3). With t-SNE and UMAP
the error is relatively uniformly distributed, but with PaCMAP there is
more error in the center of the 2D layout, reflecting that there are
more points in the middle that are further from the fitted model.

@@

\caption{\label{fig-one-dens_clust-error}The sparse end of the structure
exhibits a high model error, while the dense end shows a low model
error. The tSNE layouts of the C-shaped cluster, with uniform points
(a1) and dense points (b1), along with their selected projections (a2
and b2), are displayed. Darker colors represent a high error, while
lighter colors indicate a low error. The uniformly distributed points
demonstrate random model errors throughout the structure, whereas the
dense structure shows large model error at the sparse end.}

>>>>

To illustrate a difference in how the methods organise points in the
NLDR layout, simulated \(4\text{-}D\) data having five Gaussian clusters
is used. Figure~\ref{fig-five-gau-projs} a1, b1, c1 show the \gD{}
layouts for (a) tSNE, (b) UMAP, and (c) PaCMAP, respectively. The
default hyper-parameters are used. All three methods show the five
clusters, with varying degrees of separation.

The models are fitted to these layouts, but we focus on a single cluster
to illustrate the curious detail. Figure~\ref{fig-five-gau-projs} a2,
b2, c2 show the fitted models in a projection of the \(4\text{-}D\)
space. In this projection the difference between methods can be seen.
These clusters are fully \(4\text{-}D\) in nature, so we would expect
the model to be a \emph{crumpled \gD{} sheet} that stretches in all four
dimensions. This is what is observed for tSNE and UMAP. The curious
detail is that the model for PaCMAP is closer to a \emph{pancake} in
shape! What this means is that there has to be some ordering of points
in the \gD{} PaCMAP layout that induces the flat model, likely that the
points are organised by the global principal components. So the layout
of the model in \(4\text{-}D\) doesn't reflect the dimension of each
cluster. DHC: is there something in the documentation that might help to
explain what has happened?

This pattern can also be observed by examining the error from the model
fit (Figure~\ref{fig-five-gau-projs} a3, b3, c3). With tSNE and UMAP the
error is relatively uniformly distributed, but with PaCMAP there is more
error in the center of the \gD{} layout, reflecting that there are more
points in the middle that are further from the fitted model.

@@

\caption{\label{fig-one-dens_clust-error}The sparse end of the structure
exhibits a high model error, while the dense end shows a low model
error. The tSNE layouts of the C-shaped cluster, with uniform points
(a1) and dense points (b1) are colored according to their
high-dimensional error, along with their selected projections (a2 and
b2), are displayed. Darker colors represent a high error, while lighter
colors indicate a low error. The uniformly distributed points
demonstrate random model errors throughout the structure, whereas the
dense structure shows large model error at the sparse end.}
